---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:**

- Our group's mean for colored M&Ms is far off from the true proportion of 0.5 since we got 0.9245. 
- Different groups get different results because they are all different sample sizes and were randomly selected since it was a random handful. 
- Individual extreme results will become more normal when combining all groups, and will get closer to 0.5. 
- This relates to polls and surveys in real life since the samples weren't fair or evenly distributed with equal color categories. 

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**
The sample was 12,233 and while the population is way larger if you include the actual student body and stuff at UMD. The response rate of the survey was only 23%. According to the table, the sample was pretty closely representative of the population of over 50,000, but still not entirely representative. When looking at different racial groups, the population and sample are close, but not exact. I'm curious how they ensured a representative sample with these results. 

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
population <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(population)
```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <-population |> 
  sample_n(100)
```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
population |> summarize(mean = mean(age), median = median(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics
**A**

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

Both the means and standard deviations were similar, the mean of the sample was 48.77 while the mean of the population was 49.87. The standard deviation of the sample was 18.9 while the standard deviation of the population was 17.5. There is more variance in the population since its the entire population rather than just a sample. 

**A**:

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
