---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
WH_salaries <- read_csv(url("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv"))

# This data set includes 564 White House Employees alphabetized by last name with their statuses, salaries and titles.
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

I noticed that several of the employees have their salaries listed as "0" which will skew the data since their salaries are clearly not 0. The lowest salary is 20,000 and the highest is 251,258. There aren't many detailees in this data set but I'm curious if this would throw off the data since they are only temporary employees for the government. I've noticed there are several "assistant" job titles throughout. There's a very large spread of salaries throughout this report, which will be interesting to dive deeper into. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
  WH_salaries |>
summarize(mean_salary = mean(salary),
median_salary = median(salary))
  
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
WH_salaries |>
summarize(mean_salary = mean(salary),
median_salary = median(salary),
min_salary = min(salary), 
max_salary = max(salary))

```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean salary tells us that the average salary of those who work at the WH is 107,036, whereas the median, or middle of the data spread is $110,500. Since the median is higher than the mean, this means that the data distribution is likely negatively skewed or skewed to the left, which makes sense as to why there are a few low values or outliers. This could be the several 0 values I mentioned above. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

My best story idea is to talk to employees who make both above and below the mean salary value and write a feature piece about what they do at the WH and if their salaries align with the amount of work they do. 