---
title: "HW4 MD CityCrime"
---

In this assignment, we will be working with a dataset of total crimes reported by Maryland cities for two consecutive years. You'll load the data, calculate the violent and property crime rates for each city for both years, and then calculate the percentage change in the crime rates between the two years. You'll also write a short story based on your findings.

### Task 1: Load the Tidyverse and Janitor

You will need to load the tidyverse library and the janitor library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv. You will create a dataframe called `md_cities` that contains the data from this URL.

### Task 2: Load the Data

```{r}
md_cities <- read_csv(url("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_cities_crime_2019_2020.csv"))
```
### Task 3: Calculating Rates

Calculate violent and property crime rates for each city for both years using `mutate()`. You'll create four new columns: `violent_rate_2019`, `property_rate_2019`, `violent_rate_2020`, and `property_rate_2020`. You'll need to figure out the proper unit rate, given that there are varying populations. There's no perfect choice; choose the one that works best for you. Save the results to a new dataframe called `md_cities_rates`.

```{r}
md_cities_rates <- md_cities |>
  mutate(violent_rate_2019 = violent2019/pop2019, property_rate_2019 = property2019/pop2019, violent_rate_2020 = violent2020/pop2020, property_rate_2020 = property2020/pop2020)
```

### Task 4: Describe the Rates

Examine the new dataframe, `md_cities_rates`, by clicking on it in the Environment tab in RStudio and sorting by the rate columns. Write a couple of sentences about the results, comparing and contrasting the cities with the highest rates (both violent and property) in 2019 and 2020. Do you see any patterns? What does this tell you about crime in Maryland cities? What's the story here?


In 2019, the city with the highest violent crime rate was Baltimore with 0.0183. The city with the highest property crime rate in 2019 was Ocean City with 0.1036. In 2020, the city with the highest violent crime rate was Ocean City with 0.0239.The city with the highest property crime rate in 2020 was Ocean City with 0.1213. Ocean City seems to be taking the top spot in crime for 3/4 categories, with Baltimore having the highest crime rate in 2019. The story must be about Ocean City having a high crime rate, especially during 2019 and 2020. Could this be because people were quarantining at the beach during Covid? 


### Task 5: Calculate Percentage Change

Let's add two percentage change columns to your dataframe: one for violent crime rates and one for property crime rates. You'll calculate the percentage change between 2019 and 2020 for each city. Save the results to a new dataframe called `md_cities_change`.

```{r}
md_cities_change <- md_cities_rates |> 
  mutate(
    violent_rate_change = ((violent_rate_2020 - violent_rate_2019) / violent_rate_2019) *100,
    property_rate_change = ((property_rate_2020 - property_rate_2019) / property_rate_2019) *100
  )
```

### Task 7: Describe Percentage Changes

Look at the `md_cities_change` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change results. What's more interesting here, the patterns or the outliers? What's the story in the data?

The outliers are the most interesting in this story. Edmonston, which didn't stand out too much in the data had a percentage change of 895.32 for violent crime while Crisfield had a percentage change of 320.8078 for property crime. While Ocean City stood out to be consistent throughout with high crime rates, its percentage change for violent crime and property crime were -12.88 and -25.93 respectively. I still think the story in the data is regarding Ocean City. 


### Task 8: Tell the Best Story

Write up a lede and nutgraf of a story with the _best_ story idea you've found in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

While Baltimore has been coined the most "dangerous city in Maryland" and one of the most dangerous cities in the world, Marylanders may be shocked to find out that Ocean City had the highest crime rates in 2019 and 2020. 

At the beginning and peak of the pandemic, Ocean City had the highest property crime rates in 2019 and the highest violent and property crime rates in 2020. In 2019, Ocean City's property crime rate was 0.1036. In 2020, the city's violent crime and property crime rates were 0.0239 and 0.1213 respectively. The 2019-2020 time frame of this makes sense if more people than usual were quarantining at their beach houses in Ocean City. 

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.

### Task 10: Take the violent crime rates for 2020 and do three things:

Using the 2020 violent crime rate, calculate the percentiles using `ntile()`. Save the results to a new dataframe called `md_cities_percentiles`. How many cities are above the 90th percentile?

Calculate the Interquartile Range of the 2020 violent crime rate using `summarize()` and R's `IQR()` function. The syntax is similar to how `mean()` works.

Do the results change your best story idea from the previous assignment? Why or why not?

There are 10 cities above the 90th percentile including Baltimore and Ocean City, which makes sense. 
The results reinforce my story idea about focusing on the crime in Ocean City since this IQR reveals new information about the crime in the city that I can use for my story. 

```{r}
md_cities_percentiles <- md_cities_rates |>
  mutate(percentile_2020 = ntile (violent_rate_2020, 100))

#IQR calculation 
md_cities_percentiles %>% 
  summarise(IQR = IQR(violent_rate_2020))
```
