---
title: "HW7_MandMs"
date: 10/1/2025
---


In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```

### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mmdata <- read_csv(url("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv"))
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)
sample5 <- mmdata |> 
  sample_n(5)

sample10 <- mmdata |> 
  sample_n(10)
```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample5 |> summarize(mean = mean(red), median = median(red), sd = sd(red))
sample10 |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
mmdata |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}
mmdata |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 3) +
labs(
    title = "Total Red M&Ms data",
    x = "red",                            
    y = "Count"  )

sample5 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 3) +
labs(
    title = "Sample5 Red M&Ms data",
    x = "red",                            
    y = "Count"  )

sample10 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 3) +
labs(
    title = "Sample10 Red M&Ms data",
    x = "red",                            
    y = "Count"  )
```


### Task 7: Answer questions


Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?

The total data mean is 72.98, sample5 is 73.8	and sample10 is 74.7. The total sd is 16.45808 and sample 5 is 14.92314 and sample10 is 10.34999. The mean is more helpful to me in understanding the accuracy of the samples since they are all relatively close while the sd varies between samples and the population. 

What does this exercise tell you about the relationship between sample size and estimate accuracy?

This shows that sample sizes vary and aren't fully representative of the sample. Sample 10 has a much smaller standard deviation than sample10, which are both smaller than the total standard deviation. 

How does random sampling introduce variation into statistical estimates?

Random sampling introduces variation into statistical estimates by showing how two samples will never match perfectly and represent or mirror the entire group. No two samples are the same and can show how much variation there is in the entire population. 

If you repeated this process, would the samples always be identical? Why or why not?
No, it's almost impossible to get the same 2 samples twice since they are random. This is due to the concept of random chance and each sample will vary in what it represents. When trying to be random, you go in blind, so it's difficult to mirror the same sample twice. 


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
