---
title: "final_project"
author: "Samantha Cohen"
date: "2025-11-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

## Load data
```{r}
Billboard_Hot_100_Number_Ones_Database_Data <- read.csv("Downloads/Billboard Hot 100 Number Ones Database - Data.csv")
```

## Questions 1-3

What does the data contain and what time period does it cover?

**This data contains every song to ever top the Billboard Top 100 between August 4, 1958 and January 11, 2025. The data was compiled by author Chris Dalla Riva who wrote a book titled "Uncharted Territory: What Numbers Tell Us about the Biggest Hit Songs and Ourselves" https://bio.site/uncharted_territory. The data is organized into several categories from the basics (song title, artist, week to hit number one, how many weeks it was at number one, etc), the rating, record label, genre and style of the song assigned by Chris Dalla Riva and Vinnie Christopher, artist info (how they became famous, demographics of the artist, etc), songwriter and producer info, musical info (energy, BPM, danceability, happiness, etc) which I think will be worth looking deeper into, instrument info, structural and lyrical info, and more.**

Which columns will be most important for your analysis?

**The columns which will be most important for my analysis, at first glance, will be "Weeks at Number One," "Overall Rating," all the genre columns, "Producers" and all of the categories regarding danceability, BPM, happiness, energy. While "Weeks at Number One" and "Overall Rating" will give me a place to start looking and organizing my data, it's the other columns which cause the song's success to top the charts. Categories regarding danceability, BPM, happiness, energy and the genre and producers can be considered the "cause" columns while the "Weeks at Number One" and "Overall Rating" can be the "effect" columns. We will use them as a guide to decide what factors into making a top song.**

What's missing that you might need or want?

**Something that's missing that I might need or want for my analysis is maybe number of streams or an estimate to the number of listeners during the weeks the songs were on the top of the charts. This may be helpful to see truly how popular the songs were and which songs were fan favorites because sometimes a song will be at the top of the charts, but it won't necessarily be a fan favorite to a certain demographic. Also, demographics of the average listener per song could be helpful. Is a song more popular for the younger population? Middle-aged or older population? While the data includes a ton of different categories that I wouldn't even think of off the top of my head, there are still some that could be useful.**

## Question 4

Select a relevant column and produce basic descriptive statistics on the mean, median, lowest and highest values.

```{r}
summary_stats <- Billboard_Hot_100_Number_Ones_Database_Data |>
  summarise(
    mean_weeks = mean(`Weeks.at.Number.One`),
    median_weeks = median(`Weeks.at.Number.One`),
    min_weeks = min(`Weeks.at.Number.One`), 
    max_weeks = max(`Weeks.at.Number.One`),
    sd_weeks  = sd(`Weeks.at.Number.One`)
  )

summary_stats
```

##Newsworthy hypothesis 

Next, propose a newsworthy hypothesis (or several of them) for your dataset. You'll be looking to find (or eliminate possible story ideas) using the techniques you've learned in this class. You'll start that effort this week by coming up with some theories or questions that you're interested in exploring that can be answered with the dataset you selected.

Null: There is no relationship between danceability scores and the number of weeks a song is at number one on the charts.
 
Alternative: Songs with higher danceability scores tend to spend more weeks at number one.


I'm interested in exploring whether or not high danceability scores correlate to how many weeks a song is at the top of the charts or if the genre plays a big role too.

##Questions I have 

How will I know which is the best category to choose when investigating the dataset since there are so many categories? 


##Final Project: Part 2, Running Tests
**Run an ANOVA test**

```{r}
anova_result <- aov(Danceability ~ Weeks.at.Number.One, data = Billboard_Hot_100_Number_Ones_Database_Data)
summary(anova_result)
```
**Analysis**
According to our ANOVA test, we can see that the p-value = 0.000513, which is far below 0.05 which shows that we reject the null hypothesis, and the relationship between danceability score and weeks at number one on the charts is statistically significant. This proves that the songs with higher danceability scores are, in fact, at the top of the charts for more weeks than those with lower danceability scores. 

We must note that when looking at the strength of relationship, which can be determined by the r squared value, r^2 = 2780/(2780 + 268657) = 0.0102 shows that this model explains only about 1% of the variance. While it is statstically significant, it is still weak and the relationship between the two variables is small. 


**Run a linear regression test**
```{r}
lm_result <- lm(Danceability ~ Weeks.at.Number.One, data = Billboard_Hot_100_Number_Ones_Database_Data)
summary(lm_result)
```
**Analysis** 
This linear regression test verifies what we found in the ANOVA test. Again, we can see that the p-value = 0.000513, which is far below 0.05 which shows that we reject the null hypothesis, and the relationship between danceability score and weeks at number one on the charts is statistically significant. This proves that the songs with higher danceability scores are, in fact, at the top of the charts for more weeks than those with lower danceability scores. 

Something new we can see in this test vs. the ANOVA test is how the slope is 0.582 which means for each additional week at #1 on the charts, the danceability score increases by about 0.58 units.

Again, we can see that the strength of relationship, which can be determined by the r-squared value of 0.0102, this model explains only about 1% of the variance. While it is statistically significant, it is still weak and the relationship between the two variables is small. 

Overall, there is a statistically significant positive relationship between the weeks at number one and the charts and the danceability score although the variance is quite small. 

**Data visualization: Scatterplot**

```{r}
ggplot(Billboard_Hot_100_Number_Ones_Database_Data, 
       aes(x = Danceability, y = `Weeks.at.Number.One`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Relationship between Weeks at Number One and Danceability",
    x = "Danceability",
    y = "Weeks at Number One"
  ) 
```
**Analysis** 
We can see a weak positive relationship between Danceability and Weeks at Number One according to this scatterplot. This aligns with our r-squared value of 0.0102 which demonstrates how the relationship is weak. While of course there are outliers and some points which have low danceability scores, the overall general trend is that those with higher danceability scores are number one on the charts for more weeks. For example, the two songs which were number one on the charts for 19 weeks each, "Old Town Road" and "A Bar Song (Tipsy)" each had danceability scores of 88 and 72 respectively. While on the other hand, "Georgia on my Mind" by Ray Charles, was number one on the charts for only one week and had a danceability score of 14. These examples show the trend, but of course, there are other songs with high danceabiilty scores which were on the charts for only one week and vice versa, songs with low danceability scores which were number one on the charts for several weeks. 


